Fitness effects of mutations to SARS-CoV-2 proteins

### 2023-06-03 paper/paper.tex
revision of paper

For spike, our estimates from natural sequences correlate with the experiments almost as well as the two experimental methodologies correlate with each other (Figure~\ref{fig:dms}A), with Pearson correlations of 0.66 between the fitness estimates and experiments versus 0.72 between the two types of experiments.
Neither set of experimental measurements fully captures how mutations affect viral fitness, since both RBD yeast display and lentiviral pseudotyping are imperfect proxies for spike function during actual human infections.
Therefore, it is unclear how much the differences between the fitness estimates and experiments are due to noise in the fitness estimates versus limitations of the experiments.
However, the fact that the fitness estimates correlate with the experiments almost as well as experimental methodologies correlate with each (Figure~\ref{fig:dms}A) suggests estimates are of comparable quality to experimental measurements.
At least some of the mutations with the greatest divergence between our fitness estimates and the deep mutational scanning likely represent experimental artifacts.
For instance, P527L, which is favorable in the RBD deep mutational scan but deleterious in the sequence-based estimates and full-spike scan, is at the C-terminus of the yeast-displayed RBD~\citep{starr2020deep} where it may adopt a non-native conformation.

The sequence-based fitness estimates for Mpro also correlate with the deep mutational scans for that protein, although in this case the experiments correlate substantially better with each other than with our estimates (Figure~\ref{fig:dms}B).
However, the Mpro experiments all use a similar yeast-based methodology~\citep{flynn2022,flynn2023,iketani2022functional} that probably fails to capture significant aspects of Mpro's actual function during human infections.
For instance, a stop codon at Q306 is well tolerated in the yeast-based Mpro deep mutational scans but extremely disfavorable in our sequence-based estimates, and such a mutation would clearly be highly deleterious to actual virus as it would truncate the polyprotein.
Similarly, K61N is well tolerated in the deep mutational scans but extremely disfavorable in our estimates, probably because in the full viral polyprotein this residue mediates important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}

\subsection*{Mutation-effect estimates better capture functional constraint than dN/dS ratios or predictions from other methods}
A longstanding approach for analyzing protein constraint is to compare rates of nonsynonymous (dN) and synonymous (dS) substitutions at each site~\citep{nielsen1998likelihood,kosakovsky2005not}.
These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not}.
A known limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
Nonetheless, we can compare dN/dS ratios to our mutation-effect estimates as measures of the average constraint at each site.
The mutation-effect estimates greatly outperform dN/dS ratios as a measure of site-level constraint as assessed by correlation with deep mutational scanning experiments (Figure~\ref{fig:dnds}).
The reason is in part because some aspects of functional constraint cannot be captured by a dN/dS ratio. 
For instance, the ACE2-affinity enhancing spike mutation N501Y arose in several SARS-CoV-2 variants early in the pandemic, and has since remained fixed due its importance for receptor binding~\citep{starr2022deep}.
Our mutation-effect estimates correctly reflect that site 501 is strongly prefers tyrosine, but the site has a high dN/dS ratio due to the early convergent evolution of this site to that preferred amino acid.

Our mutation-effect estimates also correlate substantially better with deep mutational scanning experiments than predictions from two methods trained to learn mutation effects and epistatic interactions from phylogenetically broader but more sparsely sampled sequence data~\cite{thadani2022learning,rodriguez2022epistatic} (Figure~\ref{fig:comparator}).
Our estimates also correlate better with experiments than predictions of a machine-learning method that integrates a wide range of sequence and epidemiological data~\cite{maher2022predicting} (Figure~\ref{fig:comparator}).
These results suggest that our relatively straightforward approach of directly reading out the effects of mutations from their actual versus expected counts can outperform much more complex models when millions of sequences are available.

@@

To enable easy access to the mutation-effect estimates, we created an interactive plots to enable exploration of the data for each protein.
A static view of one of these plots is in Figure~\ref{fig:E_heatmap}; see \url{https://jbloomlab.github.io/SARS2-mut-fitness} for interactive versions for all proteins.
These plots enable both high-level inspection of functional constraint across each protein, and detailed interrogation of the effects of specific mutations.

@@

There are several caveats to our approach.
First, because the number of observations of any given mutation is small compared to the millions of SARS-CoV-2 sequences being analyzed, our approach requires careful quality control of publicly available sequences to remove those affected by sequencing errors.
Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{sadykov2021short,beale2004comparison}.
Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}.
Third, the exact relationship between the statistics we calculate and viral fitness depend on the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
Although we derive this relationship, we do not adjust for sampling intensity and population dynamics when estimating mutation effects.
Fourth, we make a single estimate for each mutation across all SARS-CoV-2, neglecting the epistasis that can affect some mutations~\citep{starr2022shifting,moulana2022compensatory}.
Finally there are a few technical caveats to how we count mutations that are discussed in the Methods section.

@@

\section{Methods}
\subsection{Code and data availability}
See the GitHub repository at \url{https://github.com/jbloomlab/SARS2-mut-fitness} for the computer code and processed data (eg, fitness estimates and mutation counts).
That repository contains a README with links to specific data files as well as a description of the computational pipeline.
See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aa_fitness.csv} final estimates of amino-acid fitnesses across all clades; other intermediate data files are also provided in the GitHub repository.
The specific version of the repository used for this paper is tagged as ``bioRxiv-v1'' on GitHub (\url{https://github.com/jbloomlab/SARS2-mut-fitness/tree/bioRxiv-v1})
The pipeline is fully reproducible, and is run using \texttt{snakemake}~\citep{molder2021sustainable} with interactive plots rendered using \texttt{altair}~\citep{vanderplas2018altair}.

@@

@@

We then performed quality control by ignoring any branch that met any of the following criteria:
\begin{itemize}
\item it had more than four nucleotide mutations;
\item it contained more than one nucleotide mutation that was a reversion to the Wuhan-Hu-1 reference sequence;
\item it contained more than one nucleotide mutation that was a reversion to the founder sequence for that clade as provided at \url{https://raw.githubusercontent.com/neherlab/SC2_variant_rates/7e738194a8c6592082f1caa9a6ca70cb68289790/data/clade_gts.json} by \cite{neher2022contributions};
\item it contained more than one nucleotide mutation to the same codon.
\end{itemize}
The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors, and the rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
We ignore branches with multiple nucleotide mutations to the same codon (this is very rare) because as detailed below our method is only designed to make estimates for mutations that represent single-nucleotide changes from the clade founder.
Note also that the mutation-annotated tree does not include insertion or deletion mutations, and so we only consider (and make estimates for) point mutations.

@@

First, we choose one amino acid to have a fitness value of zero, since the scale of the $f_x$ values is arbitrary and there are really only 20 unique parameters among the 21 $f_x$ values (there are 21 amino acids since we consider stops, but we only measure differences among them, not absolute values).
Typically if there was just one clade, we would set the wildtype value of $f_x = 0$ and then for mutations to all other amino acids $y$ we would simply have $f_y = \Delta f_{xy}$.
However, when there are multple clades with different founder amino acids, there is no longer a well defined "wildtype".
So we choose the most common non-stop parental amino-acid for the observed mutations and set that to zero.
In other words, we find $x$ that maximizes $\sum_y n_{xy}$ and set that $f_x$ value to zero.

@@

\subsection{Comparison to deep mutational scanning}
Deep mutational scanning data were taken from published studies~\citep{dadonaite2023pseudovirus,starr2022deep,flynn2022,iketani2022functional}, using the data at the links specified under the ``dms\_datasets'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}.
For the spike deep mutational scanning~\citep{dadonaite2023pseudovirus} we only included mutations with ``times seen'' values of at least three in the deep mutational scanning.
The RBD data~\citep{starr2022deep} include measurements for two phenotypes (ACE2 affinity and RBD expression), and one of the Mpro studies~\citep{flynn2022} includes measurements for three different phenotypes in yeast (growth, FRET, and transcription factor activity).
In both cases, Figure~\ref{fig:dms} shows the effect averaged across all phenotypes measured by the study.
For plots that break the correlations out by phenotype, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html}.

@@

\section{Acknowledgments}
We thank the UShER team for promptly answering and addressing GitHub issues related to use of this package and its pre-built mutation-annotated tree.
The work of JDB was supported in part by the NIH/NIAID under grants U19AI171399 and R01AI141707, and contract 75N93021C00015.
JDB is an Investigator of the Howard Hughes Medical Institute.

\section{Competing interests}
JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, the Vaccine Company, and Oncorus.
JDB receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.

@@

>>>>

For spike, our estimates from natural sequences correlate with the experiments almost as well as the two experimental methodologies correlate with each other (Figure~\ref{fig:dms}A), with Pearson correlations of 0.66 between the estimates and experiments versus 0.72 between the two experiments.
Neither experiment fully captures how mutations affect viral fitness, since both RBD yeast display and lentiviral pseudotyping are imperfect proxies for spike function during actual human infections.
Therefore, it is unclear how much the differences between the mutation-effect estimates and experiments are due to noise in the estimates versus limitations of the experiments.
However, the fact that the estimates correlate with the experiments almost as well as the experiments correlate with each (Figure~\ref{fig:dms}A) suggests the estimates are of comparable quality to experimental measurements.
At least some of the mutations with the greatest divergence between our estimates and the deep mutational scanning likely represent experimental artifacts.
For instance, P527L, which is favorable in the RBD deep mutational scan but deleterious in the sequence-based estimates and full-spike scan, is at the C-terminus of the yeast-displayed RBD~\citep{starr2020deep} where it may adopt a non-native conformation.

The sequence-based estimates for Mpro also correlate with the deep mutational scans for that protein, although in this case the experiments correlate substantially better with each other than with our estimates (Figure~\ref{fig:dms}B).
However, the Mpro experiments all use a similar yeast-based methodology~\citep{flynn2022,flynn2023,iketani2022functional} that fails to capture significant aspects of Mpro's function during human infections.
For instance, a stop codon at Q306 is well tolerated in the deep mutational scans but extremely disfavorable in our sequence-based estimates, and such a mutation would clearly be highly deleterious to actual virus as it would truncate the polyprotein.
Similarly, K61N is well tolerated in the deep mutational scans but extremely disfavorable in our estimates, probably because in the full viral polyprotein this residue mediates important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}

\subsection*{Mutation-effect estimates better capture functional constraint than dN/dS ratios or predictions from other methods}
A longstanding approach for analyzing protein constraint is to compare rates of nonsynonymous (dN) and synonymous (dS) substitutions at each site~\citep{nielsen1998likelihood,kosakovsky2005not}.
These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not}.
A limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
Nonetheless, we can compare dN/dS ratios to our mutation-effect estimates as measures of the average constraint at each site.
The mutation-effect estimates greatly outperform dN/dS ratios as a measure of site-level constraint as assessed by correlation with deep mutational scanning experiments (Figure~\ref{fig:dnds}).
The reason is in part because some aspects of functional constraint cannot be captured by a dN/dS ratio. 
For instance, the ACE2-affinity enhancing spike mutation N501Y arose in several SARS-CoV-2 variants early in the pandemic, and has since remained fixed due its importance for receptor binding~\citep{starr2022deep}.
Our mutation-effect estimates correctly reflect that site 501 is strongly prefers tyrosine, but the site has a high dN/dS ratio due to the early convergent evolution of this site to that preferred amino acid.

Our mutation-effect estimates also correlate better with deep mutational scanning experiments than predictions from two algorithms trained to learn epistatic models of mutation effects from phylogenetically broader but more sparsely sampled sequence data~\cite{thadani2022learning,rodriguez2022epistatic} (Figure~\ref{fig:comparator}).
Our estimates also correlate better with experiments than predictions by a machine-learning algorithm that integrates sequence and epidemiological data~\cite{maher2022predicting} (Figure~\ref{fig:comparator}).
These results suggest that our straightforward approach of directly reading out the effects of mutations from their actual versus expected counts can outperform much more complex models when millions of sequences are available.

@@

To enable easy access to the mutation-effect estimates, we created interactive plots to enable exploration of the data for each protein.
A static view of one of these plots is in Figure~\ref{fig:E_heatmap}; see \url{https://jbloomlab.github.io/SARS2-mut-fitness} for interactive versions for all proteins.
These plots enable both high-level inspection of functional constraint across each protein, and detailed interrogation of the effects of specific mutations.

@@

There are several caveats to our approach.
First, because the number of observations of any given mutation is small compared to the millions of SARS-CoV-2 sequences being analyzed, our approach requires careful quality control to remove sequencing errors.
Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{sadykov2021short,beale2004comparison}.
Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}---although this probably has only a minor effect, since our analyses show just a handful of synonymous sites are under strong selection (Figure~\ref{fig:synonymous}).
Third, the exact relationship between the statistics we calculate and viral fitness depend on the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
Although we derive this relationship, we do not adjust for sampling intensity and population dynamics when estimating mutation effects.
Fourth, we make a single estimate for each mutation across all SARS-CoV-2, neglecting the epistasis that can affect some mutations~\citep{starr2022shifting,moulana2022compensatory}.
Finally there are a few technical caveats to how we count mutations that are discussed in the Methods section.

@@

\section{Methods}
\subsection{Code and data availability}
See the GitHub repository at \url{https://github.com/jbloomlab/SARS2-mut-fitness} for the computer code and processed data (eg, fitness estimates and mutation counts).
That repository contains a README with links to specific data files as well as a description of the computational pipeline.
See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aa_fitness.csv} final estimates of amino-acid fitnesses across all clades; other intermediate data files are also provided in the GitHub repository.
The specific version of the repository used for this paper is tagged as ``bioRxiv-v2'' on GitHub (\url{https://github.com/jbloomlab/SARS2-mut-fitness/tree/bioRxiv-v2})
The pipeline is fully reproducible, and is run using \texttt{snakemake}~\citep{molder2021sustainable} with interactive plots rendered using \texttt{altair}~\citep{vanderplas2018altair}.

@@

\subsection{Versioning of analyses of different sequence sets}
The figures in this manuscript show analyses of the set of all publicly available sequences as of May-11-2023.
However, the pipeline can be run on different sequence sets.
The sequence sets on which the analysis is currently run include all sets listed under the ``mat\_trees'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}; these include the sets of all public sequences from several earlier dates (such as those available for the first version of this analysis), as well as the set of all sequences in GISAID as of March-29-2023.
A version of the results for each sequence set is provided in the GitHub repository (\url{https://github.com/jbloomlab/SARS2-mut-fitness} ) in subdirectories with names like ``results\_public\_2023-05-11'', and the index page for the interactive plots (\url{https://jbloomlab.github.io/SARS2-mut-fitness}) links at the bottom to plots for each sequence set.
The ``current\_mat'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml} specifies which sequence set is used to generate the main results that in the ``results'' subdirectory in the GitHub repository and are shown by default in the interactive plats; we anticipate periodically updating this to newer sequence sets as more sequences become available.
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/mat_aa_fitness_correlations.html} for the correlations among mutation effects estimated from the different sequence sets.

For the GISAID sequence set, we acknowledge the submitters of the sequences listed at the following URLs: \url{https://doi.org/10.55876/gis8.230403ab}, \url{https://doi.org/10.55876/gis8.230403hg}, \url{https://doi.org/10.55876/gis8.230403ht}, and \url{https://doi.org/10.55876/gis8.230403tg}.

@@

We then performed quality control by ignoring any branch that met any of the following criteria:
\begin{itemize}
\item it had more than four nucleotide mutations;
\item it contained more than one nucleotide mutation that was a reversion to the Wuhan-Hu-1 reference sequence;
\item it contained more than one nucleotide mutation that was a reversion to the founder sequence for that clade as provided at \url{https://raw.githubusercontent.com/neherlab/SC2_variant_rates/7e738194a8c6592082f1caa9a6ca70cb68289790/data/clade_gts.json} by \cite{neher2022contributions};
\item it contained more than one nucleotide mutation to the same codon.
\end{itemize}
The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors of viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~\ref{fig:nmut_counts}).
The rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
We ignore branches with multiple nucleotide mutations to the same codon (this is very rare) because as detailed below our method is only designed to make estimates for mutations that represent single-nucleotide changes from the clade founder.
Note also that the mutation-annotated tree does not include insertion or deletion mutations, and so we only consider (and make estimates for) point mutations.

@@

First, we choose one amino acid to have a fitness value of zero, since the scale of the $f_x$ values is arbitrary and there are really only 20 unique parameters among the 21 $f_x$ values (there are 21 amino acids since we consider stops, but we only measure differences among them, not absolute values).
Typically if there was just one clade, we would set the wildtype value of $f_x = 0$ and then for mutations to all other amino acids $y$ we would simply have $f_y = \Delta f_{xy}$.
However, when there are multiple clades with different founder amino acids, there is no longer a well defined "wildtype".
So we choose the most common non-stop parental amino-acid for the observed mutations and set that to zero.
In other words, we find $x$ that maximizes $\sum_y n_{xy}$ and set that $f_x$ value to zero.

@@

\subsection{Comparison to deep mutational scanning}
Deep mutational scanning data were taken from published studies~\citep{dadonaite2023pseudovirus,starr2022deep,flynn2022,flynn2023,iketani2022functional}, using the data at the links specified under the ``dms\_datasets'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}.
For the spike deep mutational scanning~\citep{dadonaite2023pseudovirus} we only included mutations with ``times seen'' values of at least three in the deep mutational scanning.
The RBD data~\citep{starr2022deep} include measurements for two phenotypes (ACE2 affinity and RBD expression), and one of the Mpro studies~\citep{flynn2022} includes measurements for three different phenotypes in yeast (growth, FRET, and transcription factor activity).
Figures~\ref{fig:dms}, \ref{fig:dnds}, and \ref{fig:comparator} shows the effect averaged across all phenotypes measured by each of these studies.
For plots that break the correlations out by phenotype, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html}.

\subsection{Comparison to dN/dS and other mutation-effect prediction algorithms}
For the comparison to the dN/dS approaches shown in Figure~\ref{fig:dnds}, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
The dN/dS ratios only provide a single number for each site, which cannot be directly compared to either the mutation-effect estimates or the deep mutational scanning, which estimate the effects of individual amino-acid mutations.
We therefore computed site-summary metrics of the mutation-effect estimates and the deep mutational scanning as the average effect of all measured amino-acid mutations at each site, excluding stop codons.
The correlations in Figure~\ref{fig:dnds} are with those site-summary metrics.

We also compared both our mutation-effect estimates and the spike deep mutational scanning measurements~\citep{dadonaite2023pseudovirus} to predictions from three other algorithms:
\begin{itemize}
\item the EpiScores reported by Maher et al (2022)~\cite{maher2022predicting},
\item the DCA mutability scores reported by Rodriguez-Rivas et al (2022)~\cite{rodriguez2022epistatic}, and
\item the EVE scores reported by Thadani et al (2023)~\cite{thadani2022learning}.
\end{itemize}
These comparisons are shown in Figure~\ref{fig:comparator}.
The Maher et al (2022) and Thadani et al (2023) studies report mutation-level predictions and so are compared directly to the deep mutational scanning our mutation-effect estimates; Rodriguez-Rivas et al (2023) report only site-level metrics and so are compared to site-summary metrics as for the dN/dS analysis.

@@

\section{Acknowledgments}
We thank Angie Hinrichs for providing the pre-built mutation-annotated trees and thank the UShER team for promptly answering and addressing GitHub issues related to use of this package and its pre-built trees.
We thank the sequence submitters to GISAID, who are listed in the tables cited in the Methods section.
The work of JDB was supported in part by the NIH/NIAID under grants U19AI171399 and R01AI141707, and contract 75N93021C00015.
JDB is an Investigator of the Howard Hughes Medical Institute.

\section{Competing interests}
JDB consults Apriori Bio, Aerium Therapeutics, Invivyd, the Vaccine Company, GSK, and Pfizer on topics related to viral evolution.
JDB receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.

@@

\begin{figure*}
\centering
\includegraphics[width=0.7\linewidth]{figs/nmut_counts.pdf}
\caption{
Number of mutations per branch on all branches of the SARS-CoV-2 tree with at least one mutation.
The analysis used here only considers mutations on branches with four or fewer mutations and so excludes the small fraction of highly mutated branches.
\label{fig:nmut_counts}
}
\end{figure*}
